---
title: "HW3-Spandan-Maaheshwari"
author: "Spandan Maaheshwari"
date: "`r Sys.Date()`"
output: pdf_document
---

Problem 1 ->

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library('ggplot2')

df_1980 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1980.csv")
df_1981 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1981.csv")
df_1982 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1982.csv")
df_1983 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1983.csv")
df_1984 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1984.csv")
df_1985 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1985.csv")
df_1986 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1986.csv")
df_1987 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1987.csv")
df_1988 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1988.csv")
df_1989 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1989.csv")
df_1990 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1990.csv")
df_1991 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1991.csv")
df_1992 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1992.csv")
df_1993 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1993.csv")
df_1994 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1994.csv")
df_1995 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1995.csv")
df_1996 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1996.csv")
df_1997 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1997.csv")
df_1998 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1998.csv")
df_1999 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_1999.csv")
df_2000 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2000.csv")
df_2001 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2001.csv")
df_2002 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2002.csv")
df_2003 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2003.csv")
df_2004 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2004.csv")
df_2005 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2005.csv")
df_2006 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2006.csv")
df_2007 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2007.csv")
df_2008 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2008.csv")
df_2009 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2009.csv")
df_2010 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2010.csv")
df_2011 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2011.csv")
df_2012 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2012.csv")
df_2013 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2013.csv")
df_2014 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2014.csv")
df_2015 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2015.csv")
df_2016 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2016.csv")
df_2017 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2017.csv")
df_2018 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2018.csv")
df_2019 = read.csv("/Users/SPANDAN/DS 5110/annual_aqi_by_county_2019.csv")

# Combining the data files for all years from 1980-2019
df_combine = dplyr::bind_rows(df_1980, df_1981, df_1982, df_1983, 
                 df_1984, df_1985, df_1986, df_1987, 
                 df_1988, df_1989, df_1990, df_1991, 
                 df_1992, df_1993, df_1994, df_1995, 
                 df_1996, df_1997, df_1998, df_1999,
                 df_2000, df_2001, df_2002, df_2003,
                 df_2004, df_2005, df_2006, df_2007,
                 df_2008, df_2009, df_2010, df_2011,
                 df_2012, df_2013, df_2014, df_2015,
                 df_2016, df_2017, df_2018, df_2019)
 
df_data_yr = select(df_combine,Year, Median.AQI)
df_data_yr <- group_by(df_data_yr,Year)
yr_summary <- summarise(df_data_yr, Median.AQI = mean(Median.AQI))
ggplot(yr_summary)+
  geom_line(mapping = aes(x=Year,y=Median.AQI),color="red",size=2)
```
Explanation for Problem 1:

It is clear from the above line plot that the worst air quality year, 1980, was when the AQI 
was at its highest. Finally, it saw a significant decline for more than ten years until 1996, 
following which it rose once more before declining once again in 2000. Another observation is that
the AQI trended downwards and decreased from 2000 to 2019 before eventually hitting its minimum
in that year.


Problem 2->

```{r}
library(ggplot2)
library(maps)

# Creating a new variable for decade

df_combine$decade <- with(df_combine, ifelse(Year>=1980 & Year<=1989, '1980-1989',
                                            ifelse(Year>=1990 & Year<=1999, '1990-1999',
                                                   ifelse(Year>=2000 & Year<=2009, '2000-2009',
                                                          ifelse(Year>=2010 & Year<=2019, '2010-2019','' )))))

x<- mutate(df_combine, region=tolower(df_combine$State))
x<- x %>% group_by(region,decade) %>% summarise(Median.AQI=mean(Median.AQI),.groups = 'drop')
states <- map_data("state")

map.df <- left_join(states,x, by=c("region"))
map.df <- map.df[order(map.df$order),]

ggplot(states, aes(x=long,y=lat,group=group)) +
  geom_polygon(data = map.df, mapping =aes(x=long,y=lat,group=group,fill=Median.AQI)) +
    scale_fill_gradientn(colours=rev(heat.colors(10)),na.value="grey90") + facet_wrap(~decade) +
      geom_path()+ coord_map()

```
Explanation for Problem 2:

It is clear from the aforementioned chloropleths that the Median AQI had dramatically 
decreased throughout each state over each decade, only remaining constant in a small number
of them. With a jump in AQI across Arizona for the decade of 2010-2019, Arizona and West 
Virginia had the highest AQI during the decade of 1980-1999, which was greatly lowered in
subsequent decades.

While the AQI dramatically decreased in states with higher levels of AQI during the 
1980–1989 decade—including California, Utah, Indiana, and Pennsylvania—it remained relatively
steady in California from the 1990s onward. Another finding is that across all decades, 
northern states had the lowest AQI.


Problem 3->

```{r}
df_country = read.csv("/Users/SPANDAN/DS 5110/ddf--gapminder--systema_globalis-master/ddf--gapminder--systema_globalis-master/ddf--entities--geo--country.csv")
df_world_4region = read.csv("/Users/SPANDAN/DS 5110/ddf--gapminder--systema_globalis-master/ddf--gapminder--systema_globalis-master/ddf--entities--geo--world_4region.csv")
df_count = inner_join(df_country, df_world_4region, by = "world_4region")

x<- select(df_count, 'country', 'world_4region', 'name.y')
x <- rename(x, Region = name.y)
x <- group_by(x,Region)
x_summary <- summarise(x, count = n())
ggplot(x_summary)+
  geom_bar(mapping = aes(count, Region, fill=Region), stat="identity") +
    labs(title = "Count of countries in each region") + 
      theme(plot.title = element_text(hjust = 0.5) ) + coord_flip()
```
Explanation for Problem 3:

Comparing the count of countries in each region using the bar plot concludes that there are 
highest no of countries in Asia continent, followed by Europe and Africa with countries in
American region being the least.


Problem 4->

```{r}

df_infmort = read.csv("/Users/SPANDAN/DS 5110/ddf--gapminder--systema_globalis-master/ddf--gapminder--systema_globalis-master/countries-etc-datapoints/ddf--datapoints--infant_mortality_rate_per_1000_births--by--geo--time.csv")
df_infmort <- rename(df_infmort, country = geo)
df_infmort = inner_join(df_infmort, df_count, by = "country")
df_infmort<- select(df_infmort, 'infant_mortality_rate_per_1000_births', 'time', 'name.y')
df_infmort <- rename(df_infmort, Region = name.y)

ggplot(df_infmort, aes(x=time,y=log(infant_mortality_rate_per_1000_births), color=Region)) +
  geom_point() + facet_wrap(~Region) + geom_smooth(method="lm", color="red")
    labs(x="Time",y="Infant_mortality_rate_per_1000_births",
      title="Infant mortality rate over time in each of the four world region") +
        theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

Explanation for Problem 4:

In all four regions, infant mortality rates have significantly decreased over time.
The highest death rates were initially seen on the European continent in the 1800s, but these 
rates eventually declined to very low levels. Another finding is that death rates in the Asian
and American regions have declined significantly while remaining significantly higher in the 
African continent.


Problem 5 ->

```{r}

df_lifeexp = read.csv("/Users/SPANDAN/DS 5110/ddf--gapminder--systema_globalis-master/ddf--gapminder--systema_globalis-master/countries-etc-datapoints/ddf--datapoints--life_expectancy_years--by--geo--time.csv")
df_lifeexp = inner_join(df_infmort, df_lifeexp, by = "time")

ggplot(df_lifeexp, aes(x=log(infant_mortality_rate_per_1000_births),y=life_expectancy_years, color=Region)) +
  geom_point() + facet_wrap(~Region)
      labs(x="Infant mortality rate per 1000 births",y="Life expectancy (years)",
        title="Relationship between infant mortality and life expectancy, by region") +
          theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
```

Explanation for Problem 5:

In each of the four regions, we see a negative relationship between life expectancy and the 
log of infant mortality rate. When infant mortality was little, the highest life expectancy
was on the continents of Europe and Asia. With mortality rate being significant the life 
expectancy was at peak across America and Africa region. Another observation is that as the
mortality rate starts increasing there is a noticeable decline in life expectancy across
all four regions.

